---
title: "NHL"
author: "Yufei Zou"
date: "2025-09-01"
output: html_document
---

```{r}
setwd("/Users/yufeizou/draft-research/NHL")

library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(mgcv)

files <- c(
  "NHL_2009_Draft_Full_Stats.csv",
  "NHL_2010_Draft_Full_Stats.csv",
  "NHL_2011_Draft_Full_Stats.csv",
  "NHL_2012_Draft_Full_Stats.csv",
  "NHL_2013_Draft_Full_Stats.csv",
  "NHL_2014_Draft_Full_Stats.csv",
  "NHL_2015_Draft_Full_Stats.csv",
  "NHL_2016_Draft_Full_Stats.csv",
  "NHL_2017_Draft_Full_Stats.csv",
  "NHL_2018_Draft_Full_Stats (2).csv"
)

#Data preparation
#"MM:SS" or "H:MM:SS" to minutes
parse_toi_to_minutes <- function(x) {
  if (is.na(x) || x == "") return(0)
  s <- as.character(x)
  parts <- str_split(s, ":", simplify = TRUE)
  nums  <- suppressWarnings(as.numeric(parts))
  nums  <- nums[!is.na(nums)]
  if (length(nums) == 2) {        # MM:SS
    return(nums[1] + nums[2] / 60)
  } else if (length(nums) == 3) { # H:MM:SS
    return(60 * nums[1] + nums[2] + nums[3] / 60)
  } else if (length(nums) == 1) { # already numeric minutes
    return(nums[1])
  } else {
    return(0)
  }
}

toi_cols <- paste0("TOIYr", 1:7)

read_one <- function(path) {
  df <- read_csv(path, show_col_types = FALSE)
  df$source_file <- basename(path)
  df
}
raw <- map(files, read_one) |> bind_rows()

stopifnot("Selection" %in% names(raw))
missing_toi <- setdiff(toi_cols, names(raw))
if (length(missing_toi) > 0) {
  stop("Missing TOI columns: ", paste(missing_toi, collapse = ", "))
}

for (c in toi_cols) {
  raw[[paste0(c, "_min")]] <- vapply(raw[[c]], parse_toi_to_minutes, numeric(1))
}
toi_min_cols <- paste0(toi_cols, "_min")

dat <- raw |>
  mutate(
    Selection   = suppressWarnings(as.integer(Selection)),
    TOI_sum_min = rowSums(across(all_of(toi_min_cols)), na.rm = TRUE)
  ) |>
  filter(!is.na(Selection), !is.na(TOI_sum_min))
```


# using GAM
```{r}
k<-40 #of basis functions
m<-2 #order
g0 <- gam(dat$TOI_sum_min ~ s(dat$Selection, bs="cr", k=k, m=m),
          data=dat, method="GCV.Cp") #cubic regression spline
plot(g0, shade=TRUE, main="Smooth effect of Selection on TOI")
```

the standard mgcv GAM plot of the smooth term for Selection predicting total TOI. The curve starts very high for top draft picks, meaning early selections contributr thr most to TOI

# Using smoothCon(), cubic regression spline
X: a design matrix X that maps coefficients to fitted values
S: penalty matrices 

```{r}
X <- dat$Selection
y <- dat$TOI_sum_min
B <- smoothCon(s(X, bs="cr", k=k, m=m),
               data=data.frame(X=X), absorb.cons=TRUE)[[1]]
xp <- B$xp
plot(g0, shade = TRUE, main = " Smooth effect of Selection on TOI")
abline(v = xp, lty = 3)
rug(xp, ticksize = 0.05, side = 3)
xp

```

```{r}
plot(g0, shade = TRUE, main = "Smooth effect of Selection on TOI")
abline(v = xp, lty = 3)

# Label every 5th xp along the top axis
idx <- seq(1, length(xp), by = 5)
axis(3, at = xp[idx], labels = round(xp[idx]), las = 2, cex.axis = 0.7)
axis(3, at = xp, labels = round(xp), las = 2, cex.axis = 0.5)

```

```{r}
k <- 40
m <- 2
X <- dat$Selection
y <- dat$TOI_sum_min
B <- smoothCon(s(X, bs="cr", k=k, m=m),
               data=data.frame(X=X), absorb.cons=FALSE)[[1]]
xp <- B$xp

Xmat <- B$X
```


```{r}
mc <- mono.con(xp,upper=FALSE)
Ain <- mc$A
bin <- mc$b
```

```{r}
q  <- ncol(Xmat)
G <- list(
  X   = Xmat,
  S   = B$S,          # pass the list as returned by smoothCon
  y   = dat$TOI_sum_min,
  w   = rep(1, nrow(dat)),
  sp  = g0$sp,      # reuse smoothing from unconstrained GAM
  C = B$C,
  p   = rep(0,q),
  Ain = Ain,
  bin = bin,
  off = 0
  
)

beta_hat <- pcls(G)
```

#LOESS
```{r}
library(ggplot2)
set.seed(1)
lo <- loess(TOI_sum_min ~ Selection,
            data = dat,
            span = 0.2,      
            degree = 1,       # local linear
            family = "gaussian",
            na.action = na.exclude,
            surface = "direct")

sel_min <- min(dat$Selection, na.rm = TRUE)
sel_max <- max(dat$Selection, na.rm = TRUE)
grid    <- data.frame(Selection = sel_min:sel_max)

pred <- grid |>
  mutate(Predicted_TOI_sum_minutes_Yrs1to7 = as.numeric(predict(lo, newdata = grid)))


p <- ggplot() +
  geom_point(data = dat, aes(Selection, TOI_sum_min), alpha = 0.25, size = 1) +
  geom_line(data = pred, aes(Selection, Predicted_TOI_sum_minutes_Yrs1to7), linewidth = 1) +
  labs(
    title = "LOESS: TOI  vs Selection\nNHL Draft 2009–2018",
    x = "Selection (overall pick number)",
    y = "Sum of TOI (minutes)"
  ) +
  theme_minimal(base_size = 12)
p
```
## Choose the best span(fraction of data points used to fit each local regression)
use k-fold cross-validation, compute MSE for each span
```{r}
spans <- seq(0.1, 0.7, by = 0.05)
set.seed(42)         
K <- 5
fold_id <- sample(rep(1:K, length.out = nrow(dat)))
#for each span, fit LOESS on training folds and predict on the held-out fold
cv_mse <- sapply(spans, function(s) {
  mse <- numeric(K)
  for (k in 1:K) {
    train <- dat[fold_id != k, ]
    test  <- dat[fold_id == k, ]
    
    fit <- loess(
      TOI_sum_min ~ Selection, data = train,
      span = s, degree = 1,
      family = "symmetric",   # robust to outliers
      na.action = na.exclude,
      surface = "direct"
    )
    
    pred <- predict(fit, newdata = test)
    mse[k] <- mean((test$TOI_sum_min - pred)^2, na.rm = TRUE)
  }
  mean(mse)
})
best_span <- spans[which.min(cv_mse)]
best_span

```

```{r}
plot(spans, cv_mse, type = "b", pch = 19,
     xlab = "Span", ylab = "Cross-validated MSE",
     main = "Choosing LOESS span by CV")
abline(v = best_span, col = "red", lty = 2)

```

## Repeat the above but have sum of TOIYr go from 1 to 7
```{r}
setwd("/Users/yufeizou/draft-research/NHL")

library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(mgcv)

files <- c(
  "NHL_2009_Draft_Full_Stats.csv",
  "NHL_2010_Draft_Full_Stats.csv",
  "NHL_2011_Draft_Full_Stats.csv",
  "NHL_2012_Draft_Full_Stats.csv",
  "NHL_2013_Draft_Full_Stats.csv",
  "NHL_2014_Draft_Full_Stats.csv",
  "NHL_2015_Draft_Full_Stats.csv",
  "NHL_2016_Draft_Full_Stats.csv",
  "NHL_2017_Draft_Full_Stats.csv",
  "NHL_2018_Draft_Full_Stats (2).csv"
)

#Data preparation
#"MM:SS" or "H:MM:SS" to minutes
parse_toi_to_minutes <- function(x) {
  if (is.na(x) || x == "") return(0)
  s <- as.character(x)
  parts <- str_split(s, ":", simplify = TRUE)
  nums  <- suppressWarnings(as.numeric(parts))
  nums  <- nums[!is.na(nums)]
  if (length(nums) == 2) {        # MM:SS
    return(nums[1] + nums[2] / 60)
  } else if (length(nums) == 3) { # H:MM:SS
    return(60 * nums[1] + nums[2] + nums[3] / 60)
  } else if (length(nums) == 1) { # already numeric minutes
    return(nums[1])
  } else {
    return(0)
  }
}

toi_cols <- paste0("TOIYr", 1:5)

read_one <- function(path) {
  df <- read_csv(path, show_col_types = FALSE)
  df$source_file <- basename(path)
  df
}
raw <- map(files, read_one) |> bind_rows()

stopifnot("Selection" %in% names(raw))
missing_toi <- setdiff(toi_cols, names(raw))
if (length(missing_toi) > 0) {
  stop("Missing TOI columns: ", paste(missing_toi, collapse = ", "))
}

for (c in toi_cols) {
  raw[[paste0(c, "_min")]] <- vapply(raw[[c]], parse_toi_to_minutes, numeric(1))
}
toi_min_cols <- paste0(toi_cols, "_min")

dat <- raw |>
  mutate(
    Selection   = suppressWarnings(as.integer(Selection)),
    TOI_sum_min = rowSums(across(all_of(toi_min_cols)), na.rm = TRUE)
  ) |>
  filter(!is.na(Selection), !is.na(TOI_sum_min))

k<-40 #of basis functions
m<-2 #order
g0 <- gam(dat$TOI_sum_min ~ s(dat$Selection, bs="cr", k=k, m=m),
          data=dat, method="GCV.Cp") #cubic regression spline
plot(g0, shade=TRUE, main="Smooth effect of Selection on TOI")

X <- dat$Selection
y <- dat$TOI_sum_min
B <- smoothCon(s(X, bs="cr", k=k, m=m),
               data=data.frame(X=X), absorb.cons=TRUE)[[1]]
xp <- B$xp
plot(g0, shade = TRUE, main = " Smooth effect of Selection on TOI(1-5)")
abline(v = xp, lty = 3)
rug(xp, ticksize = 0.05, side = 3)
xp

# Label every 5th xp along the top axis
idx <- seq(1, length(xp), by = 5)
axis(3, at = xp[idx], labels = round(xp[idx]), las = 2, cex.axis = 0.7)
axis(3, at = xp, labels = round(xp), las = 2, cex.axis = 0.5)


```

```{r}
library(dplyr)
library(ggplot2)
library(mgcv)
# --- assumes you already have `raw` and parse_toi_to_minutes() defined ---

# 1) Build summed outcomes for Yrs 1–5 and 1–7 (in minutes)
toi_cols_1to5 <- paste0("TOIYr", 1:5)
toi_cols_1to7 <- paste0("TOIYr", 1:7)

# ensure *_min columns exist (safe to re-create)
for (c in union(toi_cols_1to5, toi_cols_1to7)) {
  raw[[paste0(c, "_min")]] <- vapply(raw[[c]], parse_toi_to_minutes, numeric(1))
}

dat <- raw |>
  mutate(
    Selection = suppressWarnings(as.integer(Selection)),
    TOI_sum_min_1to5 = rowSums(across(all_of(paste0(toi_cols_1to5, "_min"))), na.rm = TRUE),
    TOI_sum_min_1to7 = rowSums(across(all_of(paste0(toi_cols_1to7, "_min"))), na.rm = TRUE)
  ) |>
  filter(!is.na(Selection), !is.na(TOI_sum_min_1to5), !is.na(TOI_sum_min_1to7))

# 2) Fit GAMs (same basis settings for comparability)
k <- 40; m <- 2
g_1to5 <- gam(TOI_sum_min_1to5 ~ s(Selection, bs = "cr", k = k, m = m),
              data = dat, method = "GCV.Cp")
g_1to7 <- gam(TOI_sum_min_1to7 ~ s(Selection, bs = "cr", k = k, m = m),
              data = dat, method = "GCV.Cp")

# 3) Make a common prediction grid over observed Selection
grid <- data.frame(Selection = seq(min(dat$Selection), max(dat$Selection)))

p_1to5 <- predict(g_1to5, newdata = grid, se.fit = TRUE)
p_1to7 <- predict(g_1to7, newdata = grid, se.fit = TRUE)

pred_df <- bind_rows(
  tibble(Selection = grid$Selection,
         fit = as.numeric(p_1to5$fit),
         se  = as.numeric(p_1to5$se.fit),
         curve = "TOI years 1–5"),
  tibble(Selection = grid$Selection,
         fit = as.numeric(p_1to7$fit),
         se  = as.numeric(p_1to7$se.fit),
         curve = "TOI years 1–7")
) |>
  mutate(lo = fit - 2*se, hi = fit + 2*se)

# (optional) show per-pick means as faint guide points
by_pick <- dat |>
  group_by(Selection) |>
  summarise(mean_1to5 = mean(TOI_sum_min_1to5, na.rm = TRUE),
            mean_1to7 = mean(TOI_sum_min_1to7, na.rm = TRUE),
            .groups = "drop") |>
  tidyr::pivot_longer(cols = starts_with("mean_"),
                      names_to = "curve",
                      values_to = "mean_val") |>
  mutate(curve = dplyr::recode(curve,
                               "mean_1to5" = "TOI years 1–5",
                               "mean_1to7" = "TOI years 1–7"))

# 4) Plot both curves with 95% bands and a legend
ggplot() +
  # optional: per-pick means
  geom_point(data = by_pick, aes(Selection, mean_val, color = curve),
             alpha = 0.25, size = 0.6) +
  # ribbons first so lines are on top
  geom_ribbon(data = pred_df, aes(x = Selection, ymin = lo, ymax = hi, fill = curve),
              alpha = 0.15, inherit.aes = FALSE) +
  geom_line(data = pred_df, aes(Selection, fit, color = curve), linewidth = 1) +
  labs(
    title = "GAM comparison: TOI (Years 1–5) vs TOI (Years 1–7) by Selection",
    x = "Selection (overall pick number)",
    y = "Sum of TOI minutes",
    color = "Curve",
    fill  = "Curve"
  ) +
  theme_minimal(base_size = 12)

```


## rescale-- make the total sum of fitted values matches
```{r}
total_by_curve <- pred_df |>
  group_by(curve) |>
  summarise(total = sum(fit, na.rm = TRUE), .groups = "drop")

total_by_curve

# Get totals
tot_1to5 <- total_by_curve$total[total_by_curve$curve == "TOI years 1–5"]
tot_1to7 <- total_by_curve$total[total_by_curve$curve == "TOI years 1–7"]

scale_factor <- tot_1to7 / tot_1to5
scale_factor

pred_df <- pred_df |>
  mutate(
    fit_scaled = ifelse(curve == "TOI years 1–5", fit * scale_factor, fit),
    se_scaled  = ifelse(curve == "TOI years 1–5", se  * scale_factor, se),
    lo_scaled  = fit_scaled - 2 * se_scaled,
    hi_scaled  = fit_scaled + 2 * se_scaled
  )
ggplot() +
  geom_ribbon(data = pred_df, aes(x = Selection, ymin = lo_scaled, ymax = hi_scaled, fill = curve),
              alpha = 0.15) +
  geom_line(data = pred_df, aes(Selection, fit_scaled, color = curve), linewidth = 1) +
  labs(
    title = "Rescaled GAM comparison: TOI (Years 1–5 vs 1–7) by Selection",
    subtitle = "Curves rescaled so their total sums match",
    x = "Selection (overall pick number)",
    y = "Rescaled Sum of TOI minutes",
    color = "Curve",
    fill  = "Curve"
  ) +
  theme_minimal(base_size = 12)

```

## use quantile regression growth curves
```{r}
library(quantregGrowth)
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Choose years 1–7 here. I
dat_q <- dat |>
  transmute(Selection = as.numeric(dat$Selection),
            TOI = dat$TOI_sum_min_1to7) |>
  filter(!is.na(Selection), !is.na(TOI))

# --- Fit median & 90th percentile with a monotone-decreasing smooth in Selection
fit_qr <- gcrq(
  TOI ~ ps(Selection, monotone = -1),   # -1 => non-increasing in Selection
  data = dat_q,
  tau  = c(0.5, 0.9)                    # median and 90th
)

# --- Predict on an integer grid of picks
sel_grid <- data.frame(Selection = seq(min(dat_q$Selection), max(dat_q$Selection)))
pred_mat <- predict(fit_qr, newdata = sel_grid)   # matrix: columns per tau
colnames(pred_mat) <- paste0("tau", c("50","90"))
pred_long <- cbind(sel_grid, as.data.frame(pred_mat)) |>
  pivot_longer(starts_with("tau"), names_to = "Quantile", values_to = "Predicted")

# --- Plot raw points + quantile curves
ggplot(dat_q, aes(Selection, TOI)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_line(data = pred_long, aes(Selection, Predicted, color = Quantile), linewidth = 1.2) +
  scale_color_manual(values = c("tau50" = "#1f77b4", "tau90" = "#ff7f0e"),
                     labels = c("Median (50th)", "90th percentile")) +
  labs(title = "Quantile Regression Growth Curves (quantregGrowth)",
       subtitle = "TOI vs Selection — monotone decreasing smooth",
       x = "Selection (overall pick number)", y = "TOI (minutes)", color = "Curve") +
  theme_minimal(base_size = 12)

```

```{r}
# install.packages("quantregGrowth")
library(quantregGrowth)
library(dplyr)
library(tidyr)
library(ggplot2)

# Choose outcome (years 1–7 here; change to _1to5 if needed)
dat_q <- dat |>
  transmute(
    Selection = as.numeric(Selection),
    TOI = TOI_sum_min_1to7
  ) |>
  filter(!is.na(Selection), !is.na(TOI))

# Fit quantile regression growth curves for τ = 0.1 ... 0.9
taus <- seq(0.1, 0.9, by = 0.1)

fit_qr <- gcrq(
  TOI ~ ps(Selection, monotone = -1),   # monotone decreasing smooth
  data = dat_q,
  tau  = taus
)

#  Predict on integer grid of selections
sel_grid <- data.frame(Selection = seq(min(dat_q$Selection), max(dat_q$Selection)))
pred_mat <- predict(fit_qr, newdata = sel_grid)

# Rename columns q10, q20, ..., q90
colnames(pred_mat) <- paste0("q", formatC(taus * 100, width = 2, flag = "0"))

# Long format for plotting
pred_long <- cbind(sel_grid, as.data.frame(pred_mat)) |>
  pivot_longer(starts_with("q"), names_to = "Quantile", values_to = "Predicted") |>
  mutate(
    tau = as.numeric(sub("q", "", Quantile)) / 100,
    Quantile = factor(Quantile, levels = paste0("q", formatC(taus * 100, width = 2, flag = "0")))
  )

# Plot: raw points + quantile curves
ggplot(dat_q, aes(Selection, TOI)) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_line(data = pred_long,
            aes(Selection, Predicted, color = Quantile),
            linewidth = 1) +
  scale_color_viridis_d(name = "Quantile (τ)") +
  labs(
    title = "Quantile Regression Growth Curves (τ = 0.1 … 0.9)",
    subtitle = "quantregGrowth::gcrq with monotone decreasing P-spline",
    x = "Selection (overall pick number)",
    y = "TOI (minutes)"
  ) +
  theme_minimal(base_size = 12)

```



